---
name: "Arman Farahani A17497672"
title: "class8"
format: pdf
---

It is important to consider scaling your data before analysis such as PCA

For example:
```{r}
head(mtcars)
```

```{r}
colMeans(mtcars)
```

```{r}
apply(mtcars, 2, sd)
```

```{r}
x <- scale(mtcars)
head(x)
```
```{r}
colMeans(x)
```

```{r}
round(colMeans(x), 2)
```

Key-point: It is usually always a good idea to scale your data prior to PCA...

## Breast Cancer Biopsy Analysis 

```{r}

```




```{r}
# Save your input data file into your Project directory
fna.data <- "https://bioboot.github.io/bimm143_S20/class-material/WisconsinCancer.csv"
```


```{r}
# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)
```

```{r}
head(wisc.df)
```
```{r}
#We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
diagnosis <- wisc.df[,1]
```

```{r}
head(wisc.data)
```

```{r}
head(diagnosis)
```

```{r}
table(diagnosis)
```

Remove this first `diagnosis` column from the dataset as I don't want to pass this to PCA etc. It is essentially the expert "answer" that we will compare our analysis results to.


## Exploratory data analysis

>Q1. How many observations are in this dataset

```{r}
dim(wisc.data)
```
569 observations

>Q2. How many of the observations have a malignant diagnosis?


```{r}
table(diagnosis)
```
212 diagnoses

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```
10 _means

## Principal Component Analysis

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
summary(wisc.pr)
```
Main "PC score plot", "PC1 vs PC2 plot"
```{r}
head(wisc.pr$x)
#plot(wisc.pr$x)
```

```{r}
attributes(wisc.pr)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], 
     col=as.factor(diagnosis))
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?



```{r}
summary(wisc.pr)
```
.4427

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3


>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7


>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```
It is difficult to understand, this is because it includes all variables within the excel that are recorded, some of which do not have correlation, and due to the shear amount, cause a saturation in the plot and makes it hard to read.


```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = as.factor(diagnosis) , 
     xlab = "PC1", ylab = "PC2")
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[,1 ], wisc.pr$x[,3], col = as.factor(diagnosis), 
     xlab = "PC1", ylab = "PC3")
```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```



>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation[,1]
```
-0.26085376

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```
5 PCs


```{r}
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

```{r}
plot(wisc.hclust)
```


>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```
Height is 19

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 10)
table(wisc.hclust.clusters, diagnosis)
```
2 is the best cluster vs diagnosis match for malignant cells, as it has the highest count of malignant cells per cluster

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust1 <- hclust(data.dist, method = "complete")
plot(wisc.hclust1)
```
```{r}
wisc.hclust2 <- hclust(data.dist, method = "single")
plot(wisc.hclust2)
```

```{r}
wisc.hclust3 <- hclust(data.dist, method = "average")
plot(wisc.hclust3)
```
```{r}
wisc.hclust4 <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust4)
```

I like ward.D2 since the organization pattern is the easiest to read and the most central